---
title: "Project 1: Spotify Track Popularity Research"
authors: "Basay Tayfun, Ethan Terrill, and Isaac Kim"
date: "2026-02-23"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
---
```{r init, include=FALSE}
library(ezids)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
options(scientific = TRUE, digits = 3)
```

```{r, echo=TRUE, include = FALSE}
library(tidyverse)
```

# Introduction

Our project examines...
**SMART Question:**
Genre is a key variable in our analysis.

# Summary of data set
```{r load_data}
spotify <- read.csv("dataset.csv", header = TRUE)
str(spotify)
```
The dataset has `r nrow(spotify)` observations and `r ncol(spotify)` variables.

names(spotify)

summary(spotify)
str(spotify)

# Key Variables
analysis_variables <- c(
  "popularity",
  "duration_ms",
  "danceability",
  "energy",
  "loudness",
  "speechiness",
  "acousticness",
  "instrumentalness",
  "liveness",
  "valence",
  "tempo",
  "explicit",
  "track_genre"
)

analysis_variables
summary(spotify[, analysis_variables])

# Genre Analysis
genre_counts <- table(spotify$track_genre)

length(genre_counts)       
unique(genre_counts)      
genre_counts      

# Missing Values

sum(is.na(spotify))

rwm <- which(!complete.cases(spotify))

for (r in rwm) {
  mis_cols <- names(spotify)[is.na(spotify[r, ])]
  
  cat("Track:", spotify$track_name[r], "\n")
  cat("Genre:", spotify$track_genre[r], "\n")
  cat("Missing Variable(s):", paste(mis_cols, collapse=", "), "\n")
}

# Avergaes per Genre

avg_popularity <- tapply(spotify$popularity, spotify$track_genre, mean)
avg_energy <- tapply(spotify$energy, spotify$track_genre, mean)
avg_danceability <- tapply(spotify$danceability, spotify$track_genre, mean)
avg_valence <- tapply(spotify$valence, spotify$track_genre, mean)
avg_loudness <- tapply(spotify$loudness, spotify$track_genre, mean)
avg_tempo <- tapply(spotify$tempo, spotify$track_genre, mean)
avg_acousticness <- tapply(spotify$acousticness, spotify$track_genre, mean)
avg_duration <- tapply(spotify$duration_ms, spotify$track_genre, mean)

genre_n <- tapply(spotify$popularity, spotify$track_genre, length)

genre_table <- data.frame(
  Genre = names(avg_popularity),
  N = as.numeric(genre_n),
  Avg_Popularity = as.numeric(avg_popularity),
  Avg_Energy = as.numeric(avg_energy),
  Avg_Danceability = as.numeric(avg_danceability),
  Avg_Valence = as.numeric(avg_valence),
  Avg_Loudness = as.numeric(avg_loudness),
  Avg_Tempo = as.numeric(avg_tempo),
  Avg_Acousticness = as.numeric(avg_acousticness),
  Avg_Duration_ms = as.numeric(avg_duration),
  row.names = NULL
)

# Now sort
genre_table <- genre_table[order(-genre_table$Avg_Popularity), ]

# Add rank
genre_table$Rank <- 1:nrow(genre_table)

# Move Rank to front
genre_table <- genre_table[, c("Rank", "Genre", "N",
                               "Avg_Popularity",
                               "Avg_Energy",
                               "Avg_Danceability",
                               "Avg_Valence",
                               "Avg_Loudness",
                               "Avg_Tempo",
                               "Avg_Acousticness",
                               "Avg_Duration_ms")]

genre_table

# Popularity Outlier Analysis per Genre

count_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR_value <- IQR(x)
  lower <- Q1 - 1.5 * IQR_value
  upper <- Q3 + 1.5 * IQR_value
  sum(x < lower | x > upper)
}

outlier_counts <- tapply(spotify$popularity,
                         spotify$track_genre,
                         count_outliers)

outlier_table <- data.frame(
  Genre = names(outlier_counts),
  Num_Popularity_Outliers = as.numeric(outlier_counts)
)

outlier_table

# Linear Regression for "Do audio features affect popularity differently across genres?"

interaction_model <- lm(
  popularity ~ danceability + energy + loudness + tempo + valence +
    acousticness + duration_ms +
    track_genre +
    danceability:track_genre +
    energy:track_genre,
  data = spotify
)

summary(interaction_model)

# Top 5 Genres

top5_genres <- genre_table$Genre[1:5]
top5_genres

results_top5 <- data.frame()

for (g in top5_genres) {
  
  sub <- spotify[spotify$track_genre == g, ]
  
  model <- lm(popularity ~ danceability + energy + loudness +
                tempo + valence + acousticness +
                duration_ms,
              data = sub)
  
  s <- summary(model)
  
  results_top5 <- rbind(results_top5,
                        data.frame(
                          Genre = g,
                          R_squared = s$r.squared,
                          Adj_R_squared = s$adj.r.squared
                        ))
  
  cat("\n========================\n")
  cat("GENRE:", g, "\n")
  print(s)
}

results_top5


```
